---
title: "DNA methylation analysis code_R"
author: "Dantong Zhu"
date: "2026-07-08"
output:
  html_document:
    css: style.css
    fig_caption: yes
    highlight: tango
    keep_md: yes
    toc: yes
    toc_float:
      collapsed: no
    toc_depth: 3
    self_contained: yes
    code_folding: hide
  word_document:
    toc: yes
    toc_depth: '3'
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      eval = TRUE)

library(ggplot2)
library(dplyr)
library(ggfortify)
library(ggpubr)
library(pheatmap)
library(tidyr)
library(readxl)
library(tidyverse)
library(stringr)
library(caret)
library(edgeR)
library(limma)
library(ComplexUpset)
library(ComplexHeatmap)
library(org.Mm.eg.db)
library(splines)
library(ggnewscale)
library(DMRcate)
library(plyr)
library(rGREAT)
library(msigdbr)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(glmnet)
library(ggrastr)
library(reshape2)
library(khroma)
library(knowYourCG)
library(SummarizedExperiment)
library(sesameData)
library(gprofiler2)
library(WGCNA)
library(lme4)
library(lmerTest)
library(ggforce)
library(ppcor)
library(lmtest)
library(variancePartition)
library(ggVennDiagram)
```

# 1. Import data; global methylation pattern and pca
```{r}
load("~/DNA methylation/discovery_cohort_DNAm_data.RData")
load("~/DNA methylation/validation_cohort_DNAm_data.RData")
load("~/DNA methylation/methy_prob_loci.RData") # prob loci
source("~/DNA methylation/DNA_methylation_functions.R") # functions for this analysis
beta2m = function(x) {log2(x/(1-x))}
methy_dat_dis_M = methy_dat_dis %>% apply(., 2, beta2m) %>% as.data.frame() #convert to Mvalues
methy_dat_val_M = methy_dat_val %>% apply(., 2, beta2m) %>% as.data.frame()
```

## 1. Frailty index & PCA
```{r}
## pca based on cpgs
methy_dat_cg_pca = prcomp(methy_dat_dis_M, scale. = TRUE) 

## variation partioning 
library(reformulas)
pc_matrix = methy_dat_cg_pca$x[, 1:121] %>% t() %>% as.data.frame()
methy_dat_dis_meta = methy_dat_dis_meta %>%
  mutate(logFI = log2(FI.Score)) %>%
  mutate(fi_resid = residuals(lm(logFI ~ `Age.at.assesment..days.`, data = methy_dat_dis_meta)),
         mouse_id = as.factor(Mouse.ID),
         sex = ifelse(Sex == "female", 1, 0))
var_partition_res = fitExtractVarPartModel(pc_matrix, 
                                           ~ fi_resid + `Age.at.assesment..days.` + sex + (1 |mouse_id),
                                           methy_dat_dis_meta)
var_partition_visual_tbl = rbind(c(sum(summary(methy_dat_cg_pca)$importance[2, 1:121]*var_partition_res[, 1]),
                                   sum(summary(methy_dat_cg_pca)$importance[2, 1:121]*var_partition_res[, 2]),
                                   sum(summary(methy_dat_cg_pca)$importance[2, 1:121]*var_partition_res[, 3]),
                                   sum(summary(methy_dat_cg_pca)$importance[2, 1:121]*var_partition_res[, 4])),
                                 var_partition_res[1:15, 1:4]) %>% as.data.frame() %>%
  dplyr::select(mouse_id, sex, `Age.at.assesment..days.`, fi_resid)
```

# 2. EWAS analysis on sex-inclusive subgroups
## 2.1. limma analysis with random effect to get DMPs
```{r}
## limma analysis, exclude T5 (only 2 samples available)
metadata_fm_excl = methy_dat_dis_meta %>% filter(Time.label != "T5")
methy_dat_fm_excl = methy_dat_dis_M[match(metadata_fm_excl$id, rownames(methy_dat_dis_M)), ]
sex = factor(metadata_fm_excl$Sex)
frail = log2(metadata_fm_excl$FI.Score)
design_fm_frail = model.matrix(~ 0 + sex + sex:frail)
colnames(design_fm_frail) = make.names(colnames(design_fm_frail))
fm_frail_fit2_res = get_fit2_res(metadata_fm_excl, design_fm_frail, methy_dat_dis_M)
## contrast 
mix_contrast = makeContrasts((sexfemale.frail + sexmale.frail)/2,
                         levels = colnames(coef(fm_frail_fit2_res)))
mix_tt = contrasts.fit(fm_frail_fit2_res, mix_contrast) %>% 
  eBayes() %>%
  topTable(., n = Inf, adjust.method = "BH", sort.by = "none")
mix_tt_dmp = mix_tt %>% filter(adj.P.Val < 0.05)
```

```{r}
## age as additive variable
sex = factor(metadata_fm_excl$Sex)
frail = log2(metadata_fm_excl$FI.Score)
age = metadata_fm_excl$`Age.at.assesment..days.`
design_fm_frail_age = model.matrix(~ 0 + sex + sex:frail + age)
colnames(design_fm_frail_age) = make.names(colnames(design_fm_frail_age))

fm_frail_fit2_res_age = get_fit2_res(metadata_fm_excl, design_fm_frail_age, methy_dat_dis_M) 
mix_contrast_age = makeContrasts((sexfemale.frail + sexmale.frail)/2,
                         levels = colnames(coef(fm_frail_fit2_res_age)))
mix_tt_age = contrasts.fit(fm_frail_fit2_res_age, mix_contrast_age) %>% 
  eBayes() %>%
  topTable(., n = Inf, adjust.method = "BH", sort.by = "none")
mix_tt_dmp_age = mix_tt_age %>% filter(adj.P.Val < 0.05)
```

## 2.2. Proportions in autosome and sex chromosomes (chi-squared test)
```{r}
## comparisons of cpg and gene location categories
mix_tt_cpg_loci = merge(mix_tt %>% mutate(IlmnID = rownames(.)),
                        cpg_loci,
                        by = "IlmnID")
mix_tt_cpg_loci_chrx = mix_tt_cpg_loci %>% filter(chromo == "chrX") # n = 10338
mix_tt_cpg_loci_chrx_dmp = mix_tt_cpg_loci_chrx %>% filter(adj.P.Val < 0.05) # n = 2184

mix_tt_cpg_loci_chry = mix_tt_cpg_loci %>% filter(chromo == "chrY") # n = 94
mix_tt_cpg_loci_chry_dmp = mix_tt_cpg_loci_chry %>% filter(adj.P.Val < 0.05) # n = 10

mix_tt_cpg_loci_chrmt = mix_tt_cpg_loci %>% filter(chromo == "chrMT") # n = 35
mix_tt_cpg_loci_chrmt_dmp = mix_tt_cpg_loci_chrmt %>% filter(adj.P.Val < 0.05) # n = 4

mix_tt_cpg_loci_autosome = mix_tt_cpg_loci %>% filter(!chromo %in% c("chrX", "chrY", "chrMT")) 
mix_tt_cpg_loci_autosome_dmp = mix_tt_cpg_loci_autosome %>% filter(adj.P.Val < 0.05)
```

```{r}
load("~/cpg_annotation_tbl.RData") 
cpg_dat = read.csv("~/MouseMethylation-12v1-0_A2.csv") # downloaded from Illumina web
colnames(cpg_annotation_tbl) = c("Name", "CpG_context", "Gene_context")
cpg_annotation_tbl = cpg_annotation_tbl %>%
  mutate(Gene_context_u = case_when(
    Gene_context == "Shore" ~ "tss_1500",
    TRUE ~ Gene_context))

cpg_dat_context_anno = merge(cpg_dat, cpg_annotation_tbl, by = "Name", all.x = TRUE) %>%
  as.data.frame() %>%
  filter(Probe_Type == "cg") %>%
  dplyr::select(IlmnID, CpG_context, Gene_context_u) %>%
  mutate(Gene_context = Gene_context_u) %>%
  dplyr::select(-Gene_context_u)

mix_all_anno = cpg_dat_context_anno %>%
  filter(IlmnID %in% mix_tt_cpg_loci$IlmnID)
mix_dmp_all_anno = cpg_dat_context_anno %>%
  filter(IlmnID %in% rownames(mix_tt_dmp))
mix_dmp_autosome_anno = cpg_dat_context_anno %>% 
  filter(IlmnID %in% mix_tt_cpg_loci_autosome_dmp$IlmnID)
mix_dmp_chrx_anno = cpg_dat_context_anno %>% 
  filter(IlmnID %in% mix_tt_cpg_loci_chrx_dmp$IlmnID)

mix_context_anno = Reduce(rbind, list(mix_all_anno %>% mutate(type = "all"),
                                      mix_dmp_all_anno %>% mutate(type = "dmp"),
                                      mix_dmp_autosome_anno %>% mutate(type = "autosome"),
                                      mix_dmp_chrx_anno %>% mutate(type = "chrx"))) %>%
                                      as.data.frame() 

mix_context_anno_cpg_dmp = mix_context_anno %>%
  dplyr::group_by(type, CpG_context) %>%
  dplyr::tally() %>%
  dplyr::mutate(prop_cpg = n/sum(n))

mix_context_anno_gene_dmp = mix_context_anno %>%
  dplyr::group_by(type, Gene_context) %>%
  dplyr::tally() %>%
  dplyr::mutate(prop_gene = n/sum(n))
```

## 2.3. kernal smoothing to get DMRs
```{r}
## plot dmrs_mix
mix_cpg_loci = methy_prob_loci %>%
    filter(IlmnID %in% colnames(methy_dat_dis_M)) %>%
    mutate(chromo = paste0("chr", CHR))
mix_cpg_fc = merge(mix_tt %>% mutate(IlmnID = rownames(.)), mix_cpg_loci, by = "IlmnID")
mix_dmr_all = get_dmr_modify_DMRcate(mix_cpg_fc) 
mix_dmr_greater5 = mix_dmr_all %>% filter(no.cpgs >= 5)
## select dmrs by Stouffer
mix_dmr = mix_dmr_all %>% filter(no.cpgs >= 5) %>%
  filter(Stouffer < 0.05)
## get cpgs within dmrs
mix_loci_dmrs = select_dmp_for_dmr(mix_dmr) # n = 12543
```

## 2.4. coMethDMR analysis, DMR level selection
```{r}
mix_loci_dmr_expr = methy_dat_dis_M %>% 
  filter(rownames(.) %in% metadata_fm_excl$id) %>%
  dplyr::select(mix_loci_dmrs$IlmnID) %>%
  mutate(id = rownames(.)) %>%
  merge(., metadata_fm_excl, by = "id") %>%
                  as.data.frame()

mix_loci_dmr_residual = lapply(2:(length(mix_loci_dmrs$IlmnID) + 1), function(x) {
  lm_fit = lm(mix_loci_dmr_expr[, x] ~ as.factor(Sex), data = mix_loci_dmr_expr)
  resid = residuals(lm_fit) %>% unname()
  return(resid)
}) %>% 
  Reduce(cbind, .) %>%
  as.data.frame() %>%
  `colnames<-`(mix_loci_dmrs$IlmnID)
rownames(mix_loci_dmr_residual) = mix_loci_dmr_expr$id
```

```{r}
mix_test_rdrop = test_rdrop(mix_loci_dmrs, mix_loci_dmr_residual)
mix_rdrop_loci_remain_v1 = mix_test_rdrop %>% filter(rdrop > 0.4)
mix_rdrop_dmr_above5 = table(mix_rdrop_loci_remain_v1$dmr) %>% as.data.frame() %>% filter(Freq >=5)
mix_rdrop_loci_remain_v2 = mix_rdrop_loci_remain_v1 %>% filter(dmr %in% mix_rdrop_dmr_above5$Var1)
```

```{r, message = FALSE, warning = FALSE}
## association by lmm, allowing random effect of CpGs
mix_loci_dmr_residual_meta = merge(mix_loci_dmr_residual %>% mutate(id = rownames(.)),
                           metadata_fm_excl, by = "id") %>% as.data.frame() %>%
  mutate(age = `Age.at.assesment..days.`,
         log2FI = log2(FI.Score))
mix_dmr_assoc_res = lapply(unique(mix_rdrop_loci_remain_v2$dmr), function(x) {
  loci_lmer = mix_rdrop_loci_remain_v2 %>% filter(dmr == x)
  expr_merged = mix_loci_dmr_residual_meta %>%
    dplyr::select(loci_lmer$IlmnID, id, Mouse.ID, age, log2FI) %>%
    melt(., id = c("id", "Mouse.ID", "age", "log2FI")) %>%
    as.data.frame()
  res1 = lmer(value ~ log2FI + age + (1 | Mouse.ID) + (log2FI | variable), data = expr_merged)
  res2 = lmer(value ~ age + (1 | Mouse.ID) + (age | variable), data = expr_merged)
  return(c(summary(res1)$coefficients[2, 5], summary(res2)$coefficients[2, 5]))
}) %>% Reduce(rbind, .) %>% as.data.frame()
rownames(mix_dmr_assoc_res) = unique(mix_rdrop_loci_remain_v2$dmr)
colnames(mix_dmr_assoc_res) = c("p_frail", "p_age")
mix_dmr_assoc_res_adj = mix_dmr_assoc_res %>%
  mutate(adjp_frail = p.adjust(p_frail, method = "fdr"),
         adjp_age = p.adjust(p_age, method = "fdr")) %>%
  mutate(cate = case_when(
    adjp_frail < 0.05 & adjp_age < 0.05 ~ "dual",
    adjp_frail < 0.05  ~ "frail",
    adjp_age < 0.05 ~ "age",
    TRUE ~ "other")) %>%
  mutate(dmr = rownames(.))
```

## 2.5. enrichment analysis for CpG of AIFI and ADFI
```{r}
mix_aidualdmr = mix_dmr_assoc_res_adj %>% filter(cate %in% c("frail", "dual"))
mix_aidualdmr_cpgs = mix_loci_dmrs[mix_loci_dmrs$dmr %in% mix_aidualdmr$dmr, ]

mix_addualdmr = mix_dmr_assoc_res_adj %>% filter(cate %in% c("age", "dual"))
mix_addualdmr_cpgs = mix_loci_dmrs[mix_loci_dmrs$dmr %in% mix_addualdmr$dmr, ]

mix_fidmr = mix_dmr_assoc_res_adj %>% filter(cate != "other")
mix_fidmr_cpgs = mix_loci_dmrs[mix_loci_dmrs$dmr %in% mix_fidmr$dmr, ]

mix_aidualdmrcpg_genes = prob_enriched_genes(mix_aidualdmr_cpgs$IlmnID)

mix_aidualdmr_path = prob_enriched_path(mix_aidualdmr_cpgs$IlmnID)
mix_addualdmr_path = prob_enriched_path(mix_addualdmr_cpgs$IlmnID)
mix_fidmr_path = prob_enriched_path(mix_fidmr_cpgs$IlmnID)
```

# 3. frailty methylation markers in females and males
## 3.1. limma analysis for DMPs
```{r}
## males 
metadata_male_only = methy_dat_dis_meta %>% filter(Sex == "male")
frail_male = log2(metadata_male_only$FI.Score)
design_male = model.matrix(~ frail_male)
colnames(design_male) = make.names(colnames(design_male))
male_dat_limma = methy_dat_dis_M %>% filter(rownames(.) %in% metadata_male_only$id)
male_dat_counts = male_dat_limma %>% t() %>% as.data.frame() 
metadata_male_only = male_dat_limma[match(metadata_male_only$id, rownames(male_dat_limma)), ]
maleonly_fit2_res = get_fit2_res(metadata_male_only, design_male, methy_dat_dis_M)

male_residuals = residuals(maleonly_fit2_res, male_dat_counts)

## females
metadata_female_only = methy_dat_dis_meta %>% filter(Sex == "female")
frail_female = log2(metadata_female_only$FI.Score)
design_female = model.matrix(~ frail_female)
colnames(design_female) = make.names(colnames(design_female))
female_dat_limma = methy_dat_dis_M %>% filter(rownames(.) %in% metadata_female_only$id)
female_dat_counts = female_dat_limma %>% t() %>% as.data.frame() 
metadata_female_only = female_dat_limma[match(metadata_female_only$id, rownames(female_dat_limma)), ]
femaleonly_fit2_res = get_fit2_res(metadata_female_only, design_female, methy_dat_dis_M)

female_residuals = residuals(femaleonly_fit2_res, female_dat_counts)

## female/male DMPs
frail_male_tt = topTable(maleonly_fit2_res, coef = 2, n = Inf, adjust.method = "BH", sort.by = "none")
frail_male_dmp = frail_male_tt %>% filter(adj.P.Val < 0.05) # nrow 69091
frail_female_tt = topTable(femaleonly_fit2_res, coef = 2, n = Inf, adjust.method = "BH", sort.by = "none")
frail_female_dmp = frail_female_tt %>% filter(adj.P.Val < 0.05) # nrow 90401
```

## 3.2 XCI analysis 
```{r}
## pipeline to identify XCI
female_dat = methy_dat_dis_meta %>% filter(Sex == "female")
male_dat = methy_dat_dis_meta %>% filter(Sex == "male")
methy_dat_cpg_chrx = methy_dat_dis %>% 
                    dplyr::select(all_of(cpg_chrx$IlmnID)) 
methy_dat_cpg_chrx_female = methy_dat_dis %>% 
                    dplyr::select(all_of(cpg_chrx$IlmnID)) %>%
                    filter(rownames(.) %in% female_dat$id)
methy_dat_cpg_chrx_male = methy_dat_dis %>% 
                    dplyr::select(all_of(cpg_chrx$IlmnID)) %>%
                    filter(rownames(.) %in% male_dat$id)
female_ave = colSums(methy_dat_cpg_chrx_female)/nrow(methy_dat_cpg_chrx_female)
male_ave = colSums(methy_dat_cpg_chrx_male)/nrow(methy_dat_cpg_chrx_male)

check_range_over = function(min_f, max_f, min_m, max_m) {
  if (min_m <= max_f & min_m >= min_f) {
    over = "yes"
  } else if (max_m <= max_f & max_m >= min_f) {
    over = "yes"
  } else {over = "no"}
  return(over)
}

methy_dat_cpg_chrx_female_range = lapply(cpg_chrx$IlmnID, function(x) {
  cpg_df = methy_dat_cpg_chrx_female[, x]
  return(c(min(cpg_df), max(cpg_df)))
}) %>%
  Reduce(rbind, .) %>%
  as.data.frame() %>%
  `colnames<-`(c("min_f", "max_f")) %>%
  mutate(IlmnID = cpg_chrx$IlmnID)
  
methy_dat_cpg_chrx_male_range = lapply(cpg_chrx$IlmnID, function(x) {
  cpg_df = methy_dat_cpg_chrx_male[, x]
  return(c(min(cpg_df), max(cpg_df)))
}) %>%
  Reduce(rbind, .) %>%
  as.data.frame() %>%
  `colnames<-`(c("min_m", "max_m")) %>%
  mutate(IlmnID = cpg_chrx$IlmnID)

range_over = sapply(1:length(cpg_chrx$IlmnID), function(x) {
  min_f = methy_dat_cpg_chrx_female_range[x, ]$min_f
  max_f = methy_dat_cpg_chrx_female_range[x, ]$max_f
  min_m = methy_dat_cpg_chrx_male_range[x, ]$min_m
  max_m = methy_dat_cpg_chrx_male_range[x, ]$max_m
  over = check_range_over(min_f, max_f, min_m, max_m)
  return(over)
})

fvsm_chrx_beta = data.frame(IlmnID = cpg_chrx$IlmnID,
                            female = female_ave,
                            male = male_ave,
                            diff = female_ave - male_ave) %>%
  merge(., cpg_chrx, by = "IlmnID") %>%
  merge(., methy_dat_cpg_chrx_female_range, by = "IlmnID") %>%
  merge(., methy_dat_cpg_chrx_male_range, by = "IlmnID") %>%
  as.data.frame() %>%
  mutate(over = range_over)

assign_Xci = function(female, male, over, diff) {
  if (female < 0.15 & male < 0.15) {
    if (over == "yes") {
      if (diff < 0.1) {
        xci = "escape"
      } else {xci = "variable escape"}
    } else {
      if (diff < 0.1) {
        xci = "escape"
      } else {xci = "XCI"}
    }
  } else {
    if (over == "yes") {
      if (diff < 0.1) {
        xci = "unclassifiable"
      } else {xci = "variable escape"}
    } else {
      if (diff < 0.1) {
        xci = "unclassifiable"
      } else {xci = "XCI"}
    }
  }
  return(xci)  
}
```

## 3.3. Female specific analysis for fi methylation makers
```{r}
female_cpg_fc = merge(frail_female_tt %>% mutate(IlmnID = rownames(.)), 
                      mix_cpg_loci, by = "IlmnID")
female_dmr_all = get_dmr_modify_DMRcate(female_cpg_fc) 
female_dmr_greater5 = female_dmr_all %>% filter(no.cpgs >= 5) %>%
  filter(chr != "chrY")

frail_female_dmr = female_dmr_greater5 %>% filter(Stouffer < 0.05)
frail_female_loci_dmrs = select_dmp_for_dmr(frail_female_dmr)
```

## 3.3. coMethyR analysis in females
```{r}
female_dmps_expr = methy_dat_dis_M %>% filter(rownames(.) %in% metadata_female_only$id) %>%
                      dplyr::select(unlist(frail_female_loci_dmrs$IlmnID))
female_dmps_expr_merged = merge(female_dmps_expr %>% mutate(id = rownames(.)), 
                metadata_female_only, by = "id") %>%
                  as.data.frame()
female_test_rdrop = test_rdrop(frail_female_loci_dmrs, female_dmps_expr)
female_rdrop_loci_remain_v1 = female_test_rdrop %>% filter(rdrop > 0.4)
female_rdrop_dmr_above5 = table(female_rdrop_loci_remain_v1$dmr) %>% as.data.frame() %>% filter(Freq >= 5)
female_rdrop_loci_remain_v2 = female_rdrop_loci_remain_v1 %>% filter(dmr %in% female_rdrop_dmr_above5$Var1)

female_loci_expr_assoc = female_dmps_expr_merged %>% as.data.frame() %>%
  mutate(age = `Age.at.assesment..days.`,
         log2FI = log2(FI.Score))

female_dmr_assoc_res = lapply(unique(female_rdrop_loci_remain_v2$dmr), function(x) {
  loci_lmer = female_rdrop_loci_remain_v2 %>% filter(dmr == x)
  expr_merged = female_loci_expr_assoc %>%
    dplyr::select(loci_lmer$IlmnID, id, Mouse.ID, age, log2FI) %>%
    melt(., id = c("id", "Mouse.ID", "age", "log2FI")) %>%
    as.data.frame()
  res1 = lmer(value ~ log2FI + age + (1 | Mouse.ID) + (log2FI | variable), data = expr_merged)
  res2 = lmer(value ~ age + (1 | Mouse.ID) + (age | variable), data = expr_merged)
  return(c(summary(res1)$coefficients[2, 5], summary(res2)$coefficients[2, 5]))
}) %>% Reduce(rbind, .) %>% as.data.frame()
rownames(female_dmr_assoc_res) = unique(female_rdrop_loci_remain_v2$dmr)
colnames(female_dmr_assoc_res) = c("p_frail", "p_age")

female_dmr_assoc_res_adj = female_dmr_assoc_res %>%
  mutate(adjp_frail = p.adjust(p_frail, method = "fdr"),
         adjp_age = p.adjust(p_age, method = "fdr")) %>%
  mutate(dmr_cate = case_when(
    adjp_frail < 0.05 & adjp_age < 0.05 ~ "dual",
    adjp_age < 0.05 ~ "age",
    adjp_frail < 0.05 ~ "frail",
    TRUE ~ "other"
  )) %>% 
  mutate(dmr = rownames(.))
```

## 3.4 female fidmr enrichment analysis
```{r}
female_dualdmrs = female_dmr_assoc_res_adj[female_dmr_assoc_res_adj$dmr_cate == "dual", ]
female_dualdmrs_cpgs = frail_female_loci_dmrs[frail_female_loci_dmrs$dmr %in% rownames(female_dualdmrs), ]
mix_aidualdmrcpg_genes = prob_enriched_genes(female_dualdmrs_cpgs$IlmnID)
female_addmrs = female_dmr_assoc_res_adj[female_dmr_assoc_res_adj$dmr_cate == "age", ]
female_addmrs_cpgs = frail_female_loci_dmrs[frail_female_loci_dmrs$dmr %in% rownames(female_addmrs), ]
female_fidmrs = female_dmr_assoc_res_adj %>% filter(dmr_cate != "other")
female_fidmrs_cpgs = frail_female_loci_dmrs[frail_female_loci_dmrs$dmr %in% female_fidmrs$dmr, ]
```

## 3.5. Male specific analysis for fi methylation makers  
```{r}
male_cpg_fc = merge(frail_male_tt %>% mutate(IlmnID = rownames(.)), 
                      mix_cpg_loci, by = "IlmnID")
male_dmr_all = get_dmr_modify_DMRcate(male_cpg_fc) 
male_dmr_greater5 = male_dmr_all %>% filter(no.cpgs >= 5)
frail_male_dmr = male_dmr_greater5 %>% filter(Stouffer < 0.05)
frail_male_loci_dmrs = select_dmp_for_dmr(frail_male_dmr)

male_loci_dmr_expr = methy_dat_dis_M %>% filter(rownames(.) %in% metadata_male_only$id) %>%
                      dplyr::select(unlist(frail_male_loci_dmrs$IlmnID))
male_loci_dmr_expr_merged = merge(male_loci_dmr_expr %>% mutate(id = rownames(.)), 
                metadata_male_only, by = "id") %>%
                  as.data.frame()
male_test_rdrop = test_rdrop(frail_male_loci_dmrs, male_loci_dmr_expr)
male_rdrop_loci_remain_v1 = male_test_rdrop %>% filter(rdrop > 0.4)
male_rdrop_dmr_above5 = table(male_rdrop_loci_remain_v1$dmr) %>% as.data.frame() %>% filter(Freq >=5)
male_rdrop_loci_remain_v2 = male_rdrop_loci_remain_v1 %>% filter(dmr %in% male_rdrop_dmr_above5$Var1)

male_loci_dmr_expr_assoc = male_loci_dmr_expr_merged %>%
  mutate(age = `Age.at.assesment..days.`,
         log2FI = log2(FI.Score))
## for each dmr, associated with Fi and age 
male_dmr_assoc_res = lapply(unique(male_rdrop_loci_remain_v2$dmr), function(x) {
  loci_lmer = male_rdrop_loci_remain_v2 %>% filter(dmr == x)
  expr_merged = male_loci_dmr_expr_assoc %>%
    dplyr::select(loci_lmer$IlmnID, id, Mouse.ID, age, log2FI) %>%
    melt(., id = c("id", "Mouse.ID", "age", "log2FI")) %>%
    as.data.frame()
  res1 = lmer(value ~ log2FI + age + (1 | Mouse.ID) + (log2FI | variable), data = expr_merged)
  res2 = lmer(value ~ age + (1 | Mouse.ID) + (age | variable), data = expr_merged)
  return(c(summary(res1)$coefficients[2, 5], summary(res2)$coefficients[2, 5]))
}) %>% Reduce(rbind, .) %>% as.data.frame()
rownames(male_dmr_assoc_res) = unique(male_rdrop_loci_remain_v2$dmr)
colnames(male_dmr_assoc_res) = c("p_frail", "p_age")
male_dmr_assoc_res_adj = male_dmr_assoc_res %>%
  mutate(adjp_frail = p.adjust(p_frail, method = "fdr"),
         adjp_age = p.adjust(p_age, method = "fdr")) %>%
  mutate(cate = case_when(
    adjp_frail < 0.05 & adjp_age < 0.05 ~ "dual",
    adjp_frail < 0.05 ~ "frail",
    adjp_age < 0.05 ~ "age",
    TRUE ~ "other")) %>%
  mutate(dmr = rownames(.))
```

```{r}
## male dmr classification
male_aidualdmrs = male_dmr_assoc_res_adj %>% filter(cate %in% c("frail", "dual"))
male_aidualdmrs_cpgs = frail_male_loci_dmrs[frail_male_loci_dmrs$dmr %in% male_aidualdmrs$dmr, ]
male_addualdmrs = male_dmr_assoc_res_adj %>% filter(cate %in% c("age", "dual"))
male_addualdmrs_cpgs = frail_male_loci_dmrs[frail_male_loci_dmrs$dmr %in% male_addualdmrs$dmr, ]
male_fidmrs = male_dmr_assoc_res_adj %>% filter(cate != "other")
male_fidmrs_cpgs = frail_male_loci_dmrs[frail_male_loci_dmrs$dmr %in% male_fidmrs$dmr, ] 
```

## 3.6 female_male associations with frailty
```{r}
methy_dat_dis_meta_update = methy_dat_dis_meta %>% as.data.frame() %>%
  mutate(age = `Age.at.assesment..days.`,
         logFI = log2(FI.Score),
         sex = as.factor(Sex),
         mouse_id = as.factor(Mouse.ID)) %>%
  cbind(methy_dat_dis_M, .) %>%
  as.data.frame()
fm_all_dmr_cpgs_combined = rbind(female_fidmrs_cpgs, male_fidmrs_cpgs) %>%
  as.data.frame()
frail_dmr_assoc_mix = lapply(fm_all_dmrs, function(x) {
  dmr_cpgs = fm_all_dmr_cpgs_combined %>% filter(dmr == x)
  fm_all_dmr_cpgs_combined_melt = methy_dat_dis_meta_update %>%
    dplyr::select(unique(dmr_cpgs$IlmnID), id, mouse_id, age, sex, logFI) %>%
    melt(., id = c("id", "mouse_id", "age", "sex", "logFI")) %>%
    as.data.frame()
  dmr_inter_assoc_mix = lmer(value ~ logFI*sex + (1|mouse_id) + (0 + logFI|mouse_id), data = fm_all_dmr_cpgs_combined_melt)
  return(dmr_inter_assoc_mix)
})

fm_all_dmrs = union(unique(female_fidmrs_cpgs$dmr), unique(male_fidmrs_cpgs$dmr))

test_inter = lapply(1:length(fm_all_dmrs), function(x) {
  return(c(fm_all_dmrs[[x]], summary(frail_dmr_assoc_mix[[x]])$coefficient[4, 5]))
}) %>% Reduce(rbind, .) %>%
  as.data.frame() %>% 
  mutate(adjp = p.adjust(V2, method = "fdr")) %>%
           filter(adjp < 0.1)

female_sig = female_specific_dmrs[female_specific_dmrs %in% test_inter$V1]
male_sig = male_specific_dmrs[male_specific_dmrs %in% test_inter$V1]
```

## 3.7. combined fm dmrs association
```{r, message = FALSE, warning = FALSE}
fidmr_assoc_f = lapply(fm_all_dmrs, function(x) {
  dmr_cpgs = fm_all_dmr_cpgs_combined %>% filter(dmr == x)
  dmr_cpgs_melt = methy_dat_dis_meta_update %>%
    filter(sex == "female") %>%
    dplyr::select(unique(dmr_cpgs$IlmnID), id, mouse_id, age, logFI) %>%
    melt(., id = c("id", "mouse_id", "age", "logFI")) %>%
    as.data.frame()
  dmr_assoc_res = lmer(value ~ logFI + (1|mouse_id) + (0 + logFI|mouse_id), data = dmr_cpgs_melt)
  return(dmr_assoc_res)
})

fidmr_assoc_m = lapply(fm_all_dmrs, function(x) {
  dmr_cpgs = fm_all_dmr_cpgs_combined %>% filter(dmr == x)
  dmr_cpgs_melt = methy_dat_dis_meta_update %>%
    filter(sex == "male") %>%
    dplyr::select(unique(dmr_cpgs$IlmnID), id, mouse_id, age, logFI) %>%
    melt(., id = c("id", "mouse_id", "age", "logFI")) %>%
    as.data.frame()
  dmr_assoc_res = lmer(value ~ logFI + (1|mouse_id) + (0 + logFI|mouse_id), data = dmr_cpgs_melt)
  return(dmr_assoc_res)
})
```

# 4. core dna methylation signatures
## 4.1. dmrs association with frailty in mixed-sex group
```{r}
fidmrs_combined = Reduce(union, list(mix_fidmr$dmr,
                                       female_fidmrs$dmr,
                                       male_fidmrs$dmr))
fidmrs_cpg_combined = Reduce(rbind, list(mix_fidmr_cpgs,
                                            female_fidmrs_cpgs,
                                            male_fidmrs_cpgs)) %>%
  as.data.frame() %>%
  distinct()
mix_unique_dmr = setdiff(mix_fidmr$dmr, union(female_fidmrs$dmr, male_fidmrs$dmr))
comb_frail_cpg_assoc = methy_dat_dis_meta_update %>%
  mutate(age = as.numeric(`Age.at.assesment..days.`)/1200) %>%
  dplyr::select(unlist(fidmrs_cpg_combined$IlmnID), sex, age, `Time.label`, mouse_id, logFI, id)
```

```{r, message = FALSE, warning = FALSE}
dmr_frail_coeff = lapply(1:925, function(x) {
  loci_lmer = fidmrs_cpg_combined %>% filter(dmr == unique(fidmrs_cpg_combined$dmr)[[x]])
  expr_merged = comb_frail_cpg_assoc %>%
    dplyr::select(loci_lmer$IlmnID, id, mouse_id, logFI, sex) %>%
    melt(., id = c("id", "mouse_id", "logFI", "sex")) %>%
    as.data.frame()
  res = lmer(logFI ~ value + as.factor(sex) + (1|mouse_id) + (1|variable), data = expr_merged)
  coef = summary(res)$coefficients
  return(c(coef[2, 1], coef[2, 5]))}) %>%
  Reduce(rbind, .) %>%
  as.data.frame()
```

# 5. DMRs association with future frailty outcomes
## 5.1 Sex-mixed analysis
```{r, message = FALSE, warning = FALSE}
## sex-mixed analysis
tp_group = data.frame(pre = c(rep("BL", 4), rep("T2", 3), rep("T3", 2), "T4"),
                      post = c("T2", "T3", "T4", "T5", "T3", "T4", "T5", "T4", "T5", "T5"))
## functions to get the dataframe for association study
comb_frail_cpg_assoc_prepost = prepost_data_merge(tp_group, comb_frail_cpg_assoc, fidmrs_cpg_combined$IlmnID)
tp_future_group = data.frame(pre = c(rep("BL", 6), rep("T2", 3), "T3"),
                             post = c("T2", "T2", "T2", "T3", "T3", "T4", "T3", "T3", "T4", "T4"),
                             future = c("T3", "T4", "T5", "T4", "T5", "T5", "T4", "T5", "T5", "T5"))
comb_frail_cpg_assoc_prepostfuture = prepostfuture_data_merge(tp_future_group, comb_frail_cpg_assoc, fidmrs_cpg_combined$IlmnID)

dmr_assoc_res = dmr_assoc_frailty_outcome(comb_frail_cpg_assoc, comb_frail_cpg_assoc_prepost,
                                          comb_frail_cpg_assoc_prepostfuture)
```

## 5.2 Sex-stratified analysis
```{r}
comb_frail_cpg_assoc_female = comb_frail_cpg_assoc %>% filter(sex == "female")
comb_frail_cpg_assoc_male = comb_frail_cpg_assoc %>% filter(sex == "male")
comb_frail_cpg_assoc_prepost_female = comb_frail_cpg_assoc_prepost %>% filter(sex.x == "female")
comb_frail_cpg_assoc_prepost_male = comb_frail_cpg_assoc_prepost %>% filter(sex.x == "male")
comb_frail_cpg_assoc_prepostfuture_female = comb_frail_cpg_assoc_prepostfuture %>% filter(sex == "female")
comb_frail_cpg_assoc_prepostfuture_male = comb_frail_cpg_assoc_prepostfuture %>% filter(sex == "male")

dmr_assoc_res_female = dmr_assoc_frailty_outcome_sex(comb_frail_cpg_assoc_female, comb_frail_cpg_assoc_prepost_female,
                                          comb_frail_cpg_assoc_prepostfuture_female)

dmr_assoc_res_male = dmr_assoc_frailty_outcome_sex(comb_frail_cpg_assoc_male, comb_frail_cpg_assoc_prepost_male,
                                          comb_frail_cpg_assoc_prepostfuture_male)
```

# 6. sensitivity test under intervention, dmr association, sex-mixed 
```{r}
methy_dat_val_comb_frail_cpg = methy_dat_val %>% ## need imputate
  dplyr::select(unlist(fidmrs_cpg_combined$IlmnID)) %>%
  mutate(id = rownames(.)) %>%
  merge(., methy_dat_val_meta, by = "id") %>%
  mutate(sex_tp = paste(Sex, Time.label)) 

methy_dat_val_impute = lapply(unique(methy_dat_val_comb_frail_cpg$sex_tp), function(x) {
  methy_dat_val_sextp = methy_dat_val_comb_frail_cpg %>%
    filter(sex_tp == x) 
  methy_dat_val_sextp_impute = methy_dat_val_sextp %>%
    dplyr::select(unlist(fidmrs_cpg_combined$IlmnID)) %>%
    base::apply(., 2, function(y) replace(as.numeric(y), is.na(y), mean(as.numeric(y), na.rm = TRUE))) %>%
      as.data.frame() %>%
    mutate(id = methy_dat_val_sextp$id)
  return(methy_dat_val_sextp_impute)
}) %>%
  Reduce(rbind, .) %>%
  as.data.frame() %>%
  `rownames<-`(.$id) %>%
  dplyr::select(-id)

methy_dat_val_M = methy_dat_val_impute %>% apply(., 2, beta2m) %>% as.data.frame()

comb_frail_cpg_assoc_val = methy_dat_val_M %>%
  dplyr::select(all_of(fidmrs_cpg_combined$IlmnID)) %>%
  mutate(id = rownames(.)) %>%
  merge(.,  methy_dat_val_meta, by = "id") %>%
                  as.data.frame() %>%
  mutate(age = as.numeric(`Age.at.assesment..days.`)/1200,
         logFI = log2(FI.Score),
         sex = ifelse(Sex == "female", 1, 0),
         mouse_id = as.factor(`Mouse.ID`)) %>%
  dplyr::select(unlist(fidmrs_cpg_combined$IlmnID), age, sex, `Time.label`, 
                mouse_id, logFI, id)

comb_frail_cpg_assoc_prepost_val = prepost_data_merge(tp_group, comb_frail_cpg_assoc_val, fidmrs_cpg_combined$IlmnID)
comb_frail_cpg_assoc_prepostfuture_val = prepostfuture_data_merge(tp_future_group,
                                                                  comb_frail_cpg_assoc_val,
                                                                  fidmrs_cpg_combined$IlmnID)
```

```{r}
## association with FI
dmr_frail_coeff_val = lapply(1:925, function(x) {
  loci_lmer = fidmrs_cpg_combined %>% filter(dmr == unique(fidmrs_cpg_combined$dmr)[[x]])
  expr_merged = comb_frail_cpg_assoc_val %>%
    dplyr::select(loci_lmer$IlmnID, id, mouse_id, logFI, sex) %>%
    melt(., id = c("id", "mouse_id", "logFI", "sex")) %>%
    as.data.frame()
  res = lmer(logFI ~ value + as.factor(sex) + (1|mouse_id) + (1|variable), data = expr_merged)
  coef = summary(res)$coefficients
  return(c(coef[2, 1], coef[2, 5]))}) %>%
  Reduce(rbind, .) %>%
  as.data.frame()

fidmr_assoc_f_val = lapply(fm_all_dmrs, function(x) {
  dmr_cpgs = fm_all_dmr_cpgs_combined %>% filter(dmr == x)
  dmr_cpgs_melt = comb_frail_cpg_assoc_val %>%
    filter(sex == "1") %>%
    dplyr::select(unique(dmr_cpgs$IlmnID), id, mouse_id, age, logFI) %>%
    melt(., id = c("id", "mouse_id", "age", "logFI")) %>%
    as.data.frame()
  dmr_assoc_res = lmer(value ~ logFI + (1|mouse_id) + (0 + logFI|mouse_id), data = dmr_cpgs_melt)
  return(dmr_assoc_res)
})

fidmr_assoc_m_val = lapply(fm_all_dmrs, function(x) {
  dmr_cpgs = fm_all_dmr_cpgs_combined %>% filter(dmr == x)
  dmr_cpgs_melt = comb_frail_cpg_assoc_val %>%
    filter(sex == "0") %>%
    dplyr::select(unique(dmr_cpgs$IlmnID), id, mouse_id, age, logFI) %>%
    melt(., id = c("id", "mouse_id", "age", "logFI")) %>%
    as.data.frame()
  dmr_assoc_res = lmer(value ~ logFI + (1|mouse_id) + (0 + logFI|mouse_id), data = dmr_cpgs_melt)
  return(dmr_assoc_res)
})
```

```{r}
## association with future FI outcomes
dmr_assoc_res_val = dmr_assoc_frailty_outcome(comb_frail_cpg_assoc_val, comb_frail_cpg_assoc_prepost_val,
                                              comb_frail_cpg_assoc_prepostfuture_val)

comb_frail_cpg_assoc_val_female = comb_frail_cpg_assoc_val %>% filter(sex == "1")
comb_frail_cpg_assoc_prepost_val_female = comb_frail_cpg_assoc_prepost_val %>% filter(sex.x == "1")
comb_frail_cpg_assoc_prepostfuture_val_female = comb_frail_cpg_assoc_prepostfuture_val %>% filter(sex == "1")
dmr_assoc_res_val_female = dmr_assoc_frailty_outcome_sex(comb_frail_cpg_assoc_val_female,
                                                         comb_frail_cpg_assoc_prepost_val_female,
                                                         comb_frail_cpg_assoc_prepostfuture_val_female)


comb_frail_cpg_assoc_val_male = comb_frail_cpg_assoc_val %>% filter(sex == "0")
comb_frail_cpg_assoc_prepost_val_male = comb_frail_cpg_assoc_prepost_val %>% filter(sex.x == "0")
comb_frail_cpg_assoc_prepostfuture_val_male = comb_frail_cpg_assoc_prepostfuture_val %>% filter(sex == "0")
dmr_assoc_res_val_male = dmr_assoc_frailty_outcome_sex(comb_frail_cpg_assoc_val_male,
                                                         comb_frail_cpg_assoc_prepost_val_male,
                                                         comb_frail_cpg_assoc_prepostfuture_val_male)
```

# 7. cpgs within dmrs variability/entropy
## 7.1 variability in sex-inclusive and -specific subgroups
```{r, warning = FALSE,  message = FALSE}
methy_dat_dis_M_excl = methy_dat_dis_meta_update %>%
  filter(Time.label != "T5") %>%
  dplyr::select(mix_fidmr_cpgs$IlmnID, logFI, Sex, mouse_id)

frail_cpg_assoc = lapply(1:3337, function(x) {
  models = lmer(methy_dat_dis_M_excl[, x] ~ methy_dat_dis_M_excl$logFI + methy_dat_dis_M_excl$Sex + (1| methy_dat_dis_M_excl$mouse_id) + (0 + methy_dat_dis_M_excl$logFI|methy_dat_dis_M_excl$mouse_id))
})

mix_fidmr_cpgs_cate = merge(mix_fidmr_cpgs, mix_fidmr, by = "dmr", all.x = TRUE) %>%
  as.data.frame()

frail_cpg_coeff_random = data.frame(IlmnID = mix_fidmr_cpgs$IlmnID,
                                    cate = mix_fidmr_cpgs_cate$cate,
                                    vc_inter = sapply(1:3337, function(y) 
                                      {vc1 = as.data.frame(VarCorr(frail_cpg_assoc[[y]]))[1, 4] 
                                      return(vc1)}),
                                    vc_slope = sapply(1:3337, function(z) 
                                      {vc2 = as.data.frame(VarCorr(frail_cpg_assoc[[z]]))[2, 4] 
                                      return(vc2)})) 
```

```{r}
methy_dat_dis_M_f = methy_dat_dis_meta_update %>%
  filter(Sex == "female") %>%
  dplyr::select(female_fidmrs_cpgs$IlmnID, logFI, mouse_id)

frail_cpg_assoc_f = lapply(1:6091, function(x) {
  models = lmer(methy_dat_dis_M_f[, x] ~ methy_dat_dis_M_f$logFI + (1| methy_dat_dis_M_f$mouse_id) + (0 + methy_dat_dis_M_f$logFI|methy_dat_dis_M_f$mouse_id))
})

female_fidmr_cpgs_cate = merge(female_fidmrs_cpgs, female_fidmrs, by = "dmr", all.x = TRUE) %>%
  as.data.frame()

frail_cpg_coeff_random_f = data.frame(IlmnID = female_fidmrs_cpgs$IlmnID,
                                    cate = female_fidmr_cpgs_cate$dmr_cate,
                                    vc_inter = sapply(1:6091, function(y) 
                                      {vc1 = as.data.frame(VarCorr(frail_cpg_assoc_f[[y]]))[1, 4] 
                                      return(vc1)}),
                                    vc_slope = sapply(1:6091, function(z) 
                                      {vc2 = as.data.frame(VarCorr(frail_cpg_assoc_f[[z]]))[2, 4] 
                                      return(vc2)})) 
```

```{r}
methy_dat_dis_M_m = methy_dat_dis_meta_update %>%
  filter(Sex == "male") %>%
  dplyr::select(male_fidmrs_cpgs$IlmnID, logFI, mouse_id)

frail_cpg_assoc_m = lapply(1:3000, function(x) {
  models = lmer(methy_dat_dis_M_m[, x] ~ methy_dat_dis_M_m$logFI + (1| methy_dat_dis_M_m$mouse_id) + (0 + methy_dat_dis_M_m$logFI|methy_dat_dis_M_m$mouse_id))
})

male_fidmr_cpgs_cate = merge(male_fidmrs_cpgs, male_fidmrs, by = "dmr", all.x = TRUE) %>%
  as.data.frame()

frail_cpg_coeff_random_m = data.frame(IlmnID = male_fidmrs_cpgs$IlmnID,
                                    cate = male_fidmr_cpgs_cate$cate,
                                    vc_inter = sapply(1:3000, function(y) 
                                      {vc1 = as.data.frame(VarCorr(frail_cpg_assoc_m[[y]]))[1, 4] 
                                      return(vc1)}),
                                    vc_slope = sapply(1:3000, function(z) 
                                      {vc2 = as.data.frame(VarCorr(frail_cpg_assoc_m[[z]]))[2, 4] 
                                      return(vc2)})) 
```

## 7.2 test of vmps
```{r}
## VMP analysis of frail-related CpGs
methy_fm_excl_resid = residuals(fm_frail_fit2_res, methy_counts_excl)
fm_excl_frail_cpg_resid = methy_fm_excl_resid[rownames(methy_fm_excl_resid) %in% mix_fidmr_cpgs$IlmnID, ]
methy_normal_test = data.frame(loci = rownames(fm_excl_frail_cpg_resid),
                               shapiro_p = sapply(1:nrow(fm_excl_frail_cpg_resid), function(x) {shapiro.test(fm_excl_frail_cpg_resid[x, ])$p.value})) %>%
  filter(shapiro_p >= 0.05)
methy_fm_excl_vmp_assoc = data.frame(loci = methy_normal_test$loci, 
                                     bp_p = sapply(1:nrow(methy_normal_test), function(x) {
                                     cpg = methy_normal_test[x, 1]
                                     lm_res = lm((methy_fm_excl_resid[cpg, ])^2 ~ frail)
                                     return(bptest(lm_res)$p.value)})) %>%
  mutate(adjp = p.adjust(bp_p, method = "BH")) %>%
  filter(adjp < 0.1)
```

```{r}
female_methy_frail_cpg = female_residuals[rownames(female_residuals) %in% female_fidmrs_cpgs$IlmnID, ]
female_methy_normal_test = data.frame(loci = rownames(female_methy_frail_cpg),
                               shapiro_p = sapply(1:nrow(female_methy_frail_cpg), function(x) {shapiro.test(female_methy_frail_cpg[x, ])$p.value})) %>%
  filter(shapiro_p >= 0.05)
female_methy_vmp_assoc = data.frame(loci = female_methy_normal_test$loci, 
                                    bp_p = sapply(1:nrow(female_methy_normal_test), function(x) {
                                    cpg = female_methy_normal_test[x, 1]
                                    lm_res = lm((female_methy_frail_cpg[cpg, ])^2 ~ frail_female)
                                       return(bptest(lm_res)$p.value)})) %>%
  mutate(adjp = p.adjust(bp_p, method = "BH")) %>% filter(adjp < 0.1)
```

```{r}
male_methy_frail_cpg = male_residuals[rownames(male_residuals) %in% male_fidmrs_cpgs$IlmnID, ]
male_methy_normal_test = data.frame(loci = rownames(male_methy_frail_cpg),
                               shapiro_p = sapply(1:nrow(male_methy_frail_cpg), function(x) {shapiro.test(male_methy_frail_cpg[x, ])$p.value})) %>%
  filter(shapiro_p >= 0.05)
male_methy_vmp_assoc = data.frame(loci = male_methy_normal_test$loci, 
                                  bp_p = sapply(1:nrow(male_methy_normal_test), function(x) {
                                  cpg = male_methy_normal_test[x, 1]
                                  lm_res = lm((male_methy_frail_cpg[cpg, ])^2 ~ frail_male)
                                       return(bptest(lm_res)$p.value)})) %>%
  mutate(adjp = p.adjust(bp_p, method = "BH")) %>% filter(adjp < 0.1)
```

## 7.3 male VMPs classification
```{r}
## SNR analysis to determine stochasticity
methy_dat_dis_meta_m_traject = methy_dat_dis_meta %>%
  filter(Sex == "male") %>%
  mutate(mouse_id = Mouse.ID) %>%
  filter(!mouse_id %in% c("52.2", "68.3", "71.4")) %>%
  mutate(logFI = log2(FI.Score))
  
methy_dat_dis_M_m_traject = methy_dat_dis_M %>%
  filter(rownames(.) %in% methy_dat_dis_meta_m_traject$id)
#vmp classification
#SNR trajectory
#SNR cloud
# exclude 52.2, 68.3, and 71.4. Only contains 1 time point
snr_ratio = sapply(male_methy_vmp_assoc$loci, function(x) {
  M_value = methy_dat_dis_M_m_traject[, x]
  adj_beta = (2^M_value)/(1 + 2^M_value)
  model = lm(adj_beta ~ methy_dat_dis_meta_m_traject$logFI)
  each_indv = methy_dat_dis_meta_m_traject %>%
    as.data.frame() %>%
    mutate(cpg = adj_beta,
           deviation = abs(unname(residuals(model))),
           fitted = unname(predict(model)))
  snr_z = sapply(unique(each_indv$mouse_id), function(y) {
    each_indv_z = each_indv %>% filter(mouse_id == y) %>%
      mutate(snr = fitted/deviation)
    return(mean(each_indv_z$snr))
  })
  each_indv_tp = each_indv %>%
    dplyr::group_by(Time.label) %>%
    dplyr::summarise(mean_M = mean(cpg))
  each_indv_update = merge(each_indv, each_indv_tp, by = "Time.label") %>%
    as.data.frame() %>%
    mutate(snr_cloud = mean_M/abs(cpg - mean_M))
  snr_cloud = mean(each_indv_update$snr_cloud)
  ratio = mean(snr_z)/snr_cloud
  return(ratio)
})
```

## 7.4. Shannon entropy of categories of dmrs
```{r}
methy_dat_dis_meta_excl = methy_dat_dis_meta %>% 
  as.data.frame() %>% 
  mutate(logFI = log2(FI.Score),
         mouse_id = as.factor(Mouse.ID)) %>%
  filter(Time.label != "T5")
methy_dat_dis_meta_excl = methy_dat_dis_meta_excl %>%
  mutate(Sex = as.factor(Sex))
methy_dat_dis_M_excl = methy_dat_dis_M %>% filter(rownames(.) %in% methy_dat_dis_meta_excl$id)

fm_ficpg_entropy = lapply(c("frail", "dual", "age"), function(x) {
  cpg_lst = mix_fidmr_cpgs_cate[mix_fidmr_cpgs_cate$cate == x, ]
  entropy = lapply(cpg_lst$IlmnID, function(y) {
    resid = residuals(lm(methy_dat_dis_M_excl[, y] ~ as.factor(methy_dat_dis_meta_excl$Sex)))
    mean = mean(methy_dat_dis_M_excl[, y])
    adj_M = mean + resid
    adj_beta = (2^adj_M)/(1 + 2^adj_M)
    adj_beta_cal = adj_beta*log2(adj_beta) + (1 - adj_beta)*log2(1 - adj_beta)
    return(adj_beta_cal)
    }) %>% 
    Reduce(cbind, .) %>%
    as.data.frame() %>%
    rowSums()
  return(entropy)
})

methy_dat_dis_meta_entropy = methy_dat_dis_meta_excl %>%
  mutate(frail_entropy = unname(-fm_ficpg_entropy[[1]]/335),
         dual_entropy = unname(-fm_ficpg_entropy[[2]]/1514),
         age_entropy = unname(-fm_ficpg_entropy[[3]]/1488))
```

```{r}
methy_dat_dis_meta_f = methy_dat_dis_meta %>% 
  as.data.frame() %>% 
  mutate(logFI = log2(FI.Score),
         mouse_id = as.factor(Mouse.ID)) %>%
  filter(Sex == "female")
methy_dat_dis_f = methy_dat_dis %>% filter(rownames(.) %in% methy_dat_dis_meta_f$id)

f_ficpg_entropy =  lapply(c("dual", "age"), function(x) {
  cpg_lst = female_fidmr_cpgs_cate[female_fidmr_cpgs_cate$dmr_cate == x, ]
  entropy = lapply(cpg_lst$IlmnID, function(y) {
    betas = methy_dat_dis_f[, y]
    beta_cal = betas*log2(betas) + (1 - betas)*log2(1 - betas)
    return(beta_cal)
    }) %>% 
    Reduce(cbind, .) %>%
    as.data.frame() %>%
    rowSums()
  return(entropy)
})

methy_dat_dis_meta_m = methy_dat_dis_meta %>% 
  as.data.frame() %>%
  filter(Sex == "male") %>%
  mutate(logFI = log2(FI.Score),
         mouse_id = as.factor(Mouse.ID))
methy_dat_dis_m = methy_dat_dis %>% filter(rownames(.) %in% methy_dat_dis_meta_m$id)

m_ficpg_entropy = lapply(c("frail", "dual", "age"), function(x) {
  cpg_lst = male_fidmr_cpgs_cate[male_fidmr_cpgs_cate$cate == x, ]
  entropy = lapply(cpg_lst$IlmnID, function(y) {
    betas = methy_dat_dis_m[, y]
    beta_cal = betas*log2(betas) + (1 - betas)*log2(1 - betas)
    return(beta_cal)
    }) %>% 
    Reduce(cbind, .) %>%
    as.data.frame() %>%
    rowSums()
  return(entropy)
})

methy_dat_dis_meta_entropy_f = methy_dat_dis_meta_f %>%
  mutate(dual_entropy = unname(-f_ficpg_entropy[[1]]/468),
         age_entropy = unname(-f_ficpg_entropy[[2]]/5623))

methy_dat_dis_meta_entropy_m = methy_dat_dis_meta_m %>%
  mutate(frail_entropy = unname(-m_ficpg_entropy[[1]]/178),
         dual_entropy = unname(-m_ficpg_entropy[[2]]/476),
         age_entropy = unname(-m_ficpg_entropy[[3]]/2346))
```

```{r}
sessionInfo()
```
